Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-04-29 sources/bmc_template_submission/bmc_article.tex
homogenizing 'phylogeny' and 'tree' use

\begin{keyword}
\kwd{gene tree}
\kwd{interoperability}
\kwd{open science}
\kwd{reproducibility}
\kwd{public database}
\kwd{DNA alignment}
\kwd{Open Tree of Life}
\kwd{otol}
\end{keyword}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data provides
the opportunity to simultaneously study
the evolutionary history of many taxa in a reproducible and continuous manner.
Here, we introduce Physcraper, a tool that establishes a data interoperability
framework for biological databases to automate data connections across biological databases.
Physcraper's main goal is to build upon published alignments and published phylogenies
to extend existing phylogenetic inferences with more data and taxa, which improve
phylogenetic reconstruction \cite{hillis1996inferring, natsidis2019phylogenomics},
time of divergence estimates \cite{schulte2013undersampling, soares2015influence},
biogeographic analyses \cite{kayaalp2017back},
as well as help in resolving phylogenetic conflict \cite{hedtke2006resolution, townsend2010optimal, natsidis2019phylogenomics}.
Physcraper updates a starting single locus alignment and corresponding tree
with public DNA data from GenBank \cite{benson2000genbank}, and links the tips in the updated trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}.

@@

The general Physcraper framework (Figure 1) consists of 4 steps: 1) identifying and
processing a tree and its underlying alignment; 2) performing a BLAST search of
DNA sequences from original alignment on GenBank, and filtering of new sequences;
3) profile-aligning new sequences to original alignment; 4) performing a phylogenetic
analysis and comparing the updated tree to existing phylogenies.

\subsection*{The inputs: a tree and an alignment}

Taxon names in the input tree must be standardized to the OpenTree taxonomy
\cite{ott3.2} using OpenTree's bulk Taxonomic Name Resolution Service
(TNRS) tool \cite{TNRStool}. Users can upload their
own phylogeny, or choose from among the 2, 950 curated phylogenies stored in OpenTree's
Phylesystem database \cite{phylesystemGithub} that also have
alignments available on the TreeBASE database \cite{treebase_website, supertreebase}.

@@

\subsection*{Tree reconstruction and comparison}

RAxML \cite{stamatakis2014raxml} is implemented to reconstruct a Maximum Likelihood
(ML) gene tree for each input alignment with default settings (GTRCAT model and
100 bootstrap replicates with default algorithm), using input tree as starting
tree for ML searches.
Bootstrap results are summarized using DendroPy's SumTrees module
\cite{sukumaran2010dendropy}.

Physcraper's main result is an updated phylogenetic hypothesis for the search taxon.
Updated and original tree are compared with Robinson-Foulds weighted and unweighted
metrics calculated with Dendropy \cite{sukumaran2010dendropy}, and with a node by node
comparison between the synthetic OpenTree and original and updated tree individually,
using OpenTree's conflict API \cite{redelings2017supertree}.

@@

We ran Physcraper on a laptop Linux computer to update an internal transcribed
spacer DNA region (ITS) alignment that was used to construct the tree from \cite{gottlieb2005molecular},
using a local GenBank database.  BLAST and RAxML analyses ran for 19hrs 45min, with bootstrap
analyses taking an additional 13hrs.
%% <!-- MTH: you probably need some details about the hardware, given the fact that you are discussing running times -->
The Gottlieb2005 tree \cite{gottlieb2005molecular} updated using Physcraper (Figure 2;
Physcraper updated tree from now on) displays all 41 distinct
taxa from the original study plus 231 new tips, contributing phylogenetic data to
84 additional \textit{Ilex} taxa. The best ML tree from the RAxML analysis is 99\%
resolved, with 25\% of nodes with bootstrap support $<$ 0.1 and 48\% nodes with
bootstrap support $>$ 0.75.
A large portion of internal branches are negligibly small, with 30 branches
$<$ 0.00001 substitution rate units, from which only 9 have a bootstrap support $>$ 0.75
(Figure 2).
As comparison with the Physcraper updated tree, the Yao2020 tree \cite{yao2020phylogeny}
also contains all 41 distinct taxa
sampled in the Gottlieb2005 tree \cite{gottlieb2005molecular},
while contributing phylogenetic data to 134 additional \textit{Ilex} taxa. From these,
67 taxa are also in the Physcraper updated tree.
While the Yao2020 tree \cite{yao2020phylogeny} was also constructed using
ITS as a marker, their GenBank data is not released yet. Hence, Physcraper was unable
to incorporate 68 taxa that are only on the Yao2020 tree because the DNA data is unavailable.
We also note that Physcraper incorporates 18 \textit{Ilex} taxa that are not in the Yao2020
tree \cite{yao2020phylogeny}. These taxa appear nested among other \textit{Ilex} species
and visual inspection of the DNA sequences suggests they are correctly assigned as \textit{Ilex}.
The ITS alignment that underlies the Yao2020 tree was constructed without any tool to
scrape GenBank \cite{yao2020phylogeny}, which could explain why Physcraper was able to
incorporate these 18 additional \textit{Ilex} taxa in the updated tree (Figure 2).

@@

To test the accuracy of Physcraper we designed a verification test.
We pruned 9 out of the 41 original tips of the Gottlieb2005 tree \cite{gottlieb2005molecular},
corresponding to a 20\% trim, excluding the outgroups. We then performed a Physcraper
run to test if we would recover the dropped tips.
We successfully recovered 6 out of 9 pruned tips in the updated tree.
Closer examination of results revealed that sequences for the 3 missing tips were correctly retrieved
with BLAST along with sequences from the other 6 tips recovered in the updated tree.
By following the GenBank accession numbers reported in the original publication
belonging to the ITS sequences of the missing tips, we observed that these three sequences
contain a 100 bp long gap of unidentified nucleotides (Ns) that is completely absent
from any of the sequences in the original alignment. This caused these three GenBank sequences
in particular to exceed Physcraper's default sequence length cutoff of 120\%, being thus
filtered and excluded from the alignment step onwards.
These sequences do appear in the Physcraper results in the list of matches from
GenBank which did not fit the sequence length cutoffs set in the configuration file.
This ``seqlen\_mismatch.txt" file includes the accession number, taxon, and sequence
length of all sequences filtered based on sequence length.

@@

Unlike phylogenetic placement approaches \cite{berger_performance_2011, matsen_pplacer_2010},
which add new taxa without modifying the input tree, Physcraper estimates all
the relationships anew in the context of the new data.
PUMPER \cite{izquierdo2014pumper} shares these conceptual strengths, but is no
longer under active development, is challenging to install and run, and has resulted
in very few phylogenetic analyses since its publication.

@@

Physcraper has the added advantage of facilitating the linkage of taxonomic information about tips
in the output phylogenies to data available in a variety of biological databases
\cite{rees2017automated}, such as
geographical locations for taxa from the GBIF \cite{gbif_secretariat_gbif_2019}.
Taxonomic links, and comparisons to existing published phylogenies in the OpenTree
data store can also help flag paralogous sequences.
Accidentally including paralogs as homologs is a risk in phylogenetic analyses,
and can be more prevalent in automated analyses than in manually curated analyses.
We provide users with several tools to try to assess homology of their aligned sequences.
The estimated gene tree itself is an evolutionarily explicit way to visualize gene
evolution, which in concert with taxonomic labelling can reveal paralogy.
OpenTree's conflict analysis tool informs the users of whether their tree contains major
conflicts with established taxonomy and any phylogenetic context they wish to compare to.
This tool also returns information on taxonomic and phylogenetic conflicts that exist
in the original input tree.
Detected conflicts may be a sign that taxonomy needs to be updated, or may be a sign
that non-homologous sequences have been included in the analysis.
These taxonomic and phylogenetic conflicts flag regions of the tree for the researcher
to more closely examine and assess homology.

>>>>

\begin{keyword}
\kwd{gene tree}
\kwd{gene phylogeny}
\kwd{multilocus}
\kwd{interoperability}
\kwd{open science}
\kwd{reproducibility}
\kwd{public database}
\kwd{DNA alignment}
\kwd{Open Tree of Life}
\kwd{otol}
\end{keyword}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
Phylogenies capture the shared history of organisms and provide key evolutionary
context for our biological observations \cite{dobzhansky1973nothing}.
Updating existing phylogenies with publicly available molecular sequence data provides
the opportunity to simultaneously study
the evolutionary history of many taxa in a reproducible and continuous manner.
Here, we introduce Physcraper, a tool that establishes a data interoperability
framework for biological databases to automate data connections across biological databases.
Physcraper's main goal is to build upon published alignments and published phylogenies
to extend existing phylogenetic inferences with more data and taxa, which improve
phylogenetic reconstruction \cite{hillis1996inferring, natsidis2019phylogenomics},
time of divergence estimates \cite{schulte2013undersampling, soares2015influence},
biogeographic analyses \cite{kayaalp2017back},
as well as help in resolving phylogenetic conflict \cite{hedtke2006resolution, townsend2010optimal, natsidis2019phylogenomics}.
Physcraper updates a starting single locus alignment and corresponding phylogeny
with public DNA data from GenBank \cite{benson2000genbank}, and links the tips in the updated trees to a unified, interoperable
taxonomic resource \cite{rees2017automated}.

@@

The general Physcraper framework (Figure 1) consists of 4 steps: 1) identifying and
processing a phylogeny and its underlying alignment; 2) performing a BLAST search of
DNA sequences from original alignment on GenBank, and filtering of new sequences;
3) profile-aligning new sequences to original alignment; 4) performing a phylogenetic
analysis and comparing the updated results to existing phylogenies.

\subsection*{The inputs: a phylogeny and an alignment}

Taxon names in the input phylogeny must be standardized to the OpenTree taxonomy
\cite{ott3.2} using OpenTree's bulk Taxonomic Name Resolution Service
(TNRS) tool \cite{TNRStool}. Users can upload their
own phylogeny, or choose from among the 2, 950 curated phylogenies stored in OpenTree's
Phylesystem database \cite{phylesystemGithub} that also have
alignments available on the TreeBASE database \cite{treebase_website, supertreebase}.

@@

\subsection*{Phylogenetic reconstruction and comparison}

RAxML \cite{stamatakis2014raxml} is implemented to reconstruct a Maximum Likelihood
(ML) gene phylogeny for each input alignment with default settings (GTRCAT model and
100 bootstrap replicates with default algorithm), using the input phylogeny as starting
tree for ML searches.
Bootstrap results are summarized using DendroPy's SumTrees module
\cite{sukumaran2010dendropy}.

Physcraper's main result is an updated phylogenetic hypothesis for the search taxon.
Updated and original phylpogeny are compared with Robinson-Foulds weighted and unweighted
metrics calculated with Dendropy \cite{sukumaran2010dendropy}, and with a node by node
comparison between the synthetic OpenTree and the original and updated phylogenies individually,
using OpenTree's conflict API \cite{redelings2017supertree}.

@@

We ran Physcraper on a laptop Linux computer to update an internal transcribed
spacer DNA region (ITS) alignment that was used to construct the phylogeny from \cite{gottlieb2005molecular},
using a local GenBank database.  BLAST and RAxML analyses ran for 19hrs 45min, with bootstrap
analyses taking an additional 13hrs.
%% <!-- MTH: you probably need some details about the hardware, given the fact that you are discussing running times -->
The Gottlieb2005 phylogeny \cite{gottlieb2005molecular} updated using Physcraper (Figure 2;
Physcraper updated phylogeny from now on) displays all 41 distinct
taxa from the original study plus 231 new tips, contributing phylogenetic data to
84 additional \textit{Ilex} taxa. The best ML phylogeny from the RAxML analysis is 99\%
resolved, with 25\% of nodes with bootstrap support $<$ 0.1 and 48\% nodes with
bootstrap support $>$ 0.75.
A large portion of internal branches are negligibly small, with 30 branches
$<$ 0.00001 substitution rate units, from which only 9 have a bootstrap support $>$ 0.75
(Figure 2).
As comparison with the Physcraper updated phylogeny, the Yao2020 phylogeny \cite{yao2020phylogeny}
also contains all 41 distinct taxa
sampled in the Gottlieb2005 phylogeny \cite{gottlieb2005molecular},
while contributing phylogenetic data to 134 additional \textit{Ilex} taxa. From these,
67 taxa are also in the Physcraper updated phylogeny.
While the Yao2020 phylogeny \cite{yao2020phylogeny} was also constructed using
ITS as a marker, their GenBank data is not released yet. Hence, Physcraper was unable
to incorporate 68 taxa that are only on the Yao2020 phylogeny because the DNA data is unavailable.
We also note that Physcraper incorporates 18 \textit{Ilex} taxa that are not in the Yao2020
phylogeny \cite{yao2020phylogeny}. These taxa appear nested among other \textit{Ilex} species
and visual inspection of the DNA sequences suggests they are correctly assigned as \textit{Ilex}.
The ITS alignment that underlies the Yao2020 phylogeny was constructed without any tool to
scrape GenBank \cite{yao2020phylogeny}, which could explain why Physcraper was able to
incorporate these 18 additional \textit{Ilex} taxa in the Physcraper updated phylogeny (Figure 2).

@@

To test the accuracy of Physcraper we designed a verification test.
We pruned 9 out of the 41 original tips of the Gottlieb2005 phylogeny \cite{gottlieb2005molecular},
corresponding to a 20\% trim, excluding the outgroups. We then performed a Physcraper
run to test if we would recover the dropped tips.
We successfully recovered 6 out of 9 pruned tips in the Physcraper updated phylogeny.
Closer examination of results revealed that sequences for the 3 missing tips were correctly retrieved
with BLAST along with sequences from the other 6 tips recovered in the updated phylogeny.
By following the GenBank accession numbers reported in the original publication
belonging to the ITS sequences of the missing tips, we observed that these three sequences
contain a 100 bp long gap of unidentified nucleotides (Ns) that is completely absent
from any of the sequences in the original alignment. This caused these three GenBank sequences
in particular to exceed Physcraper's default sequence length cutoff of 120\%, being thus
filtered and excluded from the alignment step onwards.
These sequences do appear in the Physcraper results in the list of matches from
GenBank which did not fit the sequence length cutoffs set in the configuration file.
This ``seqlen\_mismatch.txt" file includes the accession number, taxon, and sequence
length of all sequences filtered based on sequence length.

@@

Unlike phylogenetic placement approaches \cite{berger_performance_2011, matsen_pplacer_2010},
which add new taxa without modifying the input phylogeny, Physcraper estimates all
the relationships anew in the context of the new data.
PUMPER \cite{izquierdo2014pumper} shares these conceptual strengths, but is no
longer under active development, is challenging to install and run, and has resulted
in very few phylogenetic analyses since its publication.

@@

Physcraper has the added advantage of facilitating the linkage of taxonomic information about tips
in the output phylogenies to data available in a variety of biological databases
\cite{rees2017automated}, such as
geographical locations for taxa from the GBIF \cite{gbif_secretariat_gbif_2019}.
Taxonomic links, and comparisons to existing published phylogenies in the OpenTree
data store can also help flag paralogous sequences.
Accidentally including paralogs as homologs is a risk in phylogenetic analyses,
and can be more prevalent in automated analyses than in manually curated analyses.
We provide users with several tools to try to assess homology of their aligned sequences.
The estimated gene phylogeny itself is an evolutionarily explicit way to visualize gene
evolution, which in concert with taxonomic labelling can reveal paralogy.
OpenTree's conflict analysis tool informs the users of whether their phylogeny contains major
conflicts with established taxonomy and any phylogenetic context they wish to compare to.
This tool also returns information on taxonomic and phylogenetic conflicts that exist
in the original input phylogeny.
Detected conflicts may be a sign that taxonomy needs to be updated, or may be a sign
that non-homologous sequences have been included in the analysis.
These taxonomic and phylogenetic conflicts flag regions of the phylogeny for the researcher
to more closely examine and assess homology.
